Quantum Mechanics Assignment Two

### 2016-01-20 quantum/latex/a2.tex
Add changes from last nite

\usepackage{marvosym} % Required for symbols in the colored box
\usepackage{ifsym} % Required for symbols in the colored box

@@

\usepackage{fancyhdr}
\usepackage{physics}
\usepackage{amssymb}
\pagestyle{fancy}
\fancyhf{}
% Headers and footers can be added with the \lhead{} \rhead{} \lfoot{} \rfoot{} commands
% Example footer:
%\rfoot{\color{headings} {\sffamily Last update: \today}. Typeset with Xe\LaTeX}

@@

\section{9.9: QHO expectation values via integration, operator method}
In this problem we use both wave function integration and the operator method to find the
expectation values of various observables: x, p, $x^2$, $p^2$.

\textbf{a.) Ground state wave function calculations}
\begin{align*}
  <x> &= \int_{-\infty}^{\infty} \psi^*x\psi dx\\
  &= \int_{-\infty}^{\infty} \sqrt{\frac{m\omega}{\pi\hbar}}e^{-m\omega x^2 / \hbar}x dx = 0\\
\end{align*}
We don't need to perform the $<x>$ integral because it involves integrating an odd function over
an even domain, which always results in zero. This is to be expected, because the wave ground-state
wave function is symmetric across zero.\\

\begin{align*}
  <p> &= \int_{-\infty}^{\infty} \psi^* -i\hbar \frac{d}{dx}\psi dx\\
  &= \int_{-\infty}^{\infty}
  -i\sqrt{\frac{m\omega}{\pi\hbar}}(-m\omega x)e^{\frac{-m\omega x^2}{2\hbar}}dx = 0\\
\end{align*}
This integral is zero since we are again integrating an odd function over an even domain. This
makes sense because, since the spacial distribution of the wave function is equal on both sides
of the x-axis, the velocities (and hence momenta) should be equal as well.\\

\begin{align*}
  <x^2> &= \int_{-\infty}^{\infty} \psi^*x^2\psi dx\\
  &= \int_{-\infty}^{\infty} \sqrt{\frac{m\omega}{\pi\hbar}}e^{-m\omega x^2 / \hbar}x^2 dx\\
  &= 2\sqrt{\frac{m\omega}{\pi\hbar}}\int_{0}^{\infty} e^{-m\omega x^2 / \hbar}x^2 dx\\
  &= 2\sqrt{\frac{m\omega}{\pi\hbar}} *
  \frac{\sqrt{\pi}}{4(\frac{m\omega}{\hbar})^{3/2}} \hspace{2cm}\mbox{Wolfram}\\
  &= \frac{\hbar}{2m\omega}\\
\end{align*}
Here we exploit the fact that integrating an even function over an even domain is equivalent to
doubling the integral of the function over half the domain. The answer makes sense, since increasing
$m\omega$ corresponds to making the quadratic potential steeper, hence making the wave function
tighter.\\

\begin{align*}
  <p^2> &= \int_{-\infty}^{\infty} \psi^* \left(-\hbar^2 \frac{d^2}{dx^2}\right)\psi dx\\
  &= \sqrt{\frac{m\omega}{\pi\hbar}}\int_{-\infty}^{\infty} -\hbar^2e^{\frac{-m\omega x^2}{\hbar}}
  \left(\frac{m^2\omega^2x^2}{\hbar^2} - \frac{m\omega x}{\hbar}\right)dx\\
  &= -\hbar^2\sqrt{\frac{m\omega}{\pi\hbar}}\left(
  \int_{-\infty}^{\infty} e^{\frac{-m\omega x^2}{\hbar}}\frac{m^2\omega^2x^2}{\hbar^2}dx -
  \int_{-\infty}^{\infty} e^{\frac{-m\omega x^2}{\hbar}}\frac{m\omega}{\hbar}\right)dx\\
  &= -\hbar^2\sqrt{\frac{m\omega}{\pi\hbar}}
  \left(\frac{1}{4}\sqrt{\frac{m\pi\omega}{\hbar}} -
  \frac{1}{2}\sqrt{\frac{m\pi\omega}{\hbar}}\right) \hspace{2cm}\mbox{Wolfram}\\
  &= \frac{\hbar m\omega}{4}\\
\end{align*}
This answer is intuitive, since increasing mass as well as oscillatory speed should both increase
the expected momentum.

\textbf{b.) Any-state operator method calculations}
Here we define the position and momentum operators in terms of the raising and lowering operators:
\begin{align*}
  \hat{x} &= \sqrt{\frac{\hbar}{2m\omega}}\left(a + a^\dagger\right)\\
  \hat{p} &= i\sqrt{\frac{m\omega\hbar}{2}}\left(a^\dagger - a\right)\\
\end{align*}
We then solve for the various expectation values using these definitions:

\begin{align*}
  <x> &= \bra{n}x\ket{n}\\
  &= \sqrt{\frac{\hbar}{2m\omega}}\bra{n}\left(a + a^\dagger\right)\ket{n}\\
  &= \sqrt{\frac{\hbar}{2m\omega}}
  \left(\sqrt{n}\bra{n}\ket{n-1} + \sqrt{n+1}\bra{n}\ket{n+1}\right) = 0\\
\end{align*}
This answer indicates that all pure QHO wave functions are centered around zero. This makes
sense since the QHO wave functions are all made up of either even or odd functions. When squared,
as is done when finding the probability, the equation becomes entirely even, meaning it is
symmetric across the y-axis. From this we can see that all QHO wave functions have an expectation
value of zero for x-measurement.

\begin{align*}
  <p> &= \bra{n}\hat{p}\ket{n}\\
  &= i\sqrt{\frac{m\hbar\omega}{2}}\bra{n}\left(a^\dagger-a\right)\ket{n}\\
  &= i\sqrt{\frac{m\hbar\omega}{2}}
  \left(\sqrt{n+1}\bra{n}\ket{n+1} - \sqrt{n}\bra{n}\ket{n-1}\right) = 0\\
\end{align*}
This is an interesting result. It is intuitive that any wave function should have an expectation of
zero for momentum, else the wave function would move through space.

How does this reconcile w/ fact that p involves a spacial derivative?

>>>>

\usepackage{marvosym} % Required for symbols in the colored box

@@

\usepackage{fancyhdr}
\usepackage{amsmath}
% \usepackage{physics}
\usepackage{amssymb}
\pagestyle{fancy}
\fancyhf{}
% Headers and footers can be added with the \lhead{} \rhead{} \lfoot{} \rfoot{} commands
% Example footer:
%\rfoot{\color{headings} {\sffamily Last update: \today}. Typeset with Xe\LaTeX}

@@

%% \section{9.9: QHO expectation values via integration, operator method}
%% In this problem we use both wave function integration and the operator method to find the
%% expectation values of various observables: x, p, $x^2$, $p^2$.

%% \textbf{a.) Ground state wave function calculations}
%% \begin{align*}
%%   <x> &= \int_{-\infty}^{\infty} \psi^*x\psi dx\\
%%   &= \int_{-\infty}^{\infty} \sqrt{\frac{m\omega}{\pi\hbar}}e^{-m\omega x^2 / \hbar}x dx = 0\\
%% \end{align*}
%% We don't need to perform the $<x>$ integral because it involves integrating an odd function over
%% an even domain, which always results in zero. This is to be expected, because the wave ground-state
%% wave function is symmetric across zero.\\

%% \begin{align*}
%%   <p> &= \int_{-\infty}^{\infty} \psi^* -i\hbar \frac{d}{dx}\psi dx\\
%%   &= \int_{-\infty}^{\infty}
%%   -i\sqrt{\frac{m\omega}{\pi\hbar}}(-m\omega x)e^{\frac{-m\omega x^2}{2\hbar}}dx = 0\\
%% \end{align*}
%% This integral is zero since we are again integrating an odd function over an even domain. This
%% makes sense because, since the spacial distribution of the wave function is equal on both sides
%% of the x-axis, the velocities (and hence momenta) should be equal as well.\\

%% \begin{align*}
%%   <x^2> &= \int_{-\infty}^{\infty} \psi^*x^2\psi dx\\
%%   &= \int_{-\infty}^{\infty} \sqrt{\frac{m\omega}{\pi\hbar}}e^{-m\omega x^2 / \hbar}x^2 dx\\
%%   &= 2\sqrt{\frac{m\omega}{\pi\hbar}}\int_{0}^{\infty} e^{-m\omega x^2 / \hbar}x^2 dx\\
%%   &= 2\sqrt{\frac{m\omega}{\pi\hbar}} *
%%   \frac{\sqrt{\pi}}{4(\frac{m\omega}{\hbar})^{3/2}} \hspace{2cm}\mbox{Wolfram}\\
%%   &= \frac{\hbar}{2m\omega}\\
%% \end{align*}
%% Here we exploit the fact that integrating an even function over an even domain is equivalent to
%% doubling the integral of the function over half the domain. The answer makes sense, since increasing
%% $m\omega$ corresponds to making the quadratic potential steeper, hence making the wave function
%% tighter.\\

%% \begin{align*}
%%   <p^2> &= \int_{-\infty}^{\infty} \psi^* \left(-\hbar^2 \frac{d^2}{dx^2}\right)\psi dx\\
%%   &= \sqrt{\frac{m\omega}{\pi\hbar}}\int_{-\infty}^{\infty} -\hbar^2e^{\frac{-m\omega x^2}{\hbar}}
%%   \left(\frac{m^2\omega^2x^2}{\hbar^2} - \frac{m\omega}{\hbar}\right)dx\\
%%   &= -\hbar^2\sqrt{\frac{m\omega}{\pi\hbar}}\left(
%%   \int_{-\infty}^{\infty} e^{\frac{-m\omega x^2}{\hbar}}\frac{m^2\omega^2x^2}{\hbar^2}dx -
%%   \int_{-\infty}^{\infty} e^{\frac{-m\omega x^2}{\hbar}}\frac{m\omega}{\hbar}\right)dx\\
%%   &= -\hbar^2\sqrt{\frac{m\omega}{\pi\hbar}}
%%   \left(\frac{1}{2}\sqrt{\frac{m\pi\omega}{\hbar}} -
%%   \sqrt{\frac{m\pi\omega}{\hbar}}\right) \hspace{2cm}\mbox{Wolfram}\\
%%   &= \frac{\hbar m\omega}{2}\\
%% \end{align*}
%% This answer is intuitive, since increasing mass as well as oscillatory speed should both increase
%% the expected momentum.

%% \textbf{b.) Any-state operator method calculations}
%% Here we define the position and momentum operators in terms of the raising and lowering operators:
%% \begin{align*}
%%   \hat{x} &= \sqrt{\frac{\hbar}{2m\omega}}\left(a + a^\dagger\right)\\
%%   \hat{p} &= i\sqrt{\frac{m\omega\hbar}{2}}\left(a^\dagger - a\right)\\
%% \end{align*}
%% We then solve for the various expectation values using these definitions:

%% \begin{align*}
%%   <x> &= \bra{n}x\ket{n}\\
%%   &= \sqrt{\frac{\hbar}{2m\omega}}\bra{n}\left(a + a^\dagger\right)\ket{n}\\
%%   &= \sqrt{\frac{\hbar}{2m\omega}}
%%   \left(\sqrt{n}\bra{n}\ket{n-1} + \sqrt{n+1}\bra{n}\ket{n+1}\right) = 0\\
%% \end{align*}
%% This answer indicates that all pure QHO wave functions are centered around zero. This makes
%% sense since the QHO wave functions are all made up of either even or odd functions. When squared,
%% as is done when finding the probability, the equation becomes entirely even, meaning it is
%% symmetric across the y-axis. From this we can see that all QHO wave functions have an expectation
%% value of zero for x-measurement.

%% \begin{align*}
%%   <p> &= \bra{n}\hat{p}\ket{n}\\
%%   &= i\sqrt{\frac{m\hbar\omega}{2}}\bra{n}\left(a^\dagger-a\right)\ket{n}\\
%%   &= i\sqrt{\frac{m\hbar\omega}{2}}
%%   \left(\sqrt{n+1}\bra{n}\ket{n+1} - \sqrt{n}\bra{n}\ket{n-1}\right) = 0\\
%% \end{align*}
%% This is an interesting result. It is intuitive that any wave function should have an expectation of
%% zero for momentum, else the wave function would move through space due to a nonzero average
%% velocity.

%% \begin{align*}
%%   <x^2> &= \bra{n}x^2\ket{n}\\
%%   &= \frac{\hbar}{2m\omega}\bra{n}\left(a + a^\dagger\right)\left(a + a^\dagger\right)\ket{n}\\
%%   &= \frac{\hbar}{2m\omega}\bra{n}
%%   \left(aa + aa^\dagger + a^\dagger a + a^\dagger a^\dagger\right)\ket{n}\\
%%   &= \frac{\hbar}{2m\omega}
%%   \left(aa + aa^\dagger + a^\dagger a + a^\dagger a^\dagger\right)\ket{n}\\
%%   &= \frac{\hbar}{2m\omega} \left(\sqrt{n-1}\sqrt{n-2}\bra{n}\ket{n-2} + (n+1)\bra{n}\ket{n} +
%%   (n)\bra{n}\ket{n} + \sqrt{n+1}\sqrt{n+2}\bra{n}\ket{n+2}\right)\\
%%   &= \frac{\hbar}{2m\omega} \left(2n+1\right)\\
%% \end{align*}
%% This is an interesting result which indicates that for each extra rung on the ladder, the
%% expectation of the squared position increases linearly, meaning the expectation of the magnitude
%% of x increases as $\sqrt{n}$. This makes sense, since increasing energy levels have less and less
%% room to expand in the quadratic potential well. The $n=0$ state corresponds to the wave function
%% solution for the ground state found above.\\

%% \begin{align*}
%%   <p^2> &= \bra{n}p^2\ket{n}\\
%%   &= -\frac{m\hbar\omega}{2}\bra{n}\left(a^\dagger-a\right)\left(a^\dagger-a\right)\ket{n}\\
%%   &= -\frac{m\hbar\omega}{2}\bra{n}
%%   \left(a^\dagger a^\dagger - a^\dagger a - aa^\dagger + aa\right)\ket{n}\\
%%   &= -\frac{m\hbar\omega}{2}\bra{n}\left(\sqrt{n-1}\sqrt{n-2}\bra{n}\ket{n-2}
%%   - (n)\bra{n}\ket{n} - (n+1)\bra{n}\ket{n} + \sqrt{n+1}\sqrt{n+2}\bra{n}\ket{n+2}\right)\\
%%   &= \frac{\hbar m\omega}{2}\left(2n + 1\right)
%% \end{align*}
%% This result is similar to the above one for $x^2$ with different units. It is still a $\sqrt{n}$
%% relationship for momentum, which makes sense.\\

%% \textbf{c.) Uncertainty principle}\\
%% Since the above two cases agree with each other, we check that the more general case of part
%% \textbf{b.} agrees with the uncertainty principle.

%% \begin{align*}
%%   \Delta x &= \sqrt{<x^2>-<x>^2}\\
%%   &= \sqrt{\frac{\hbar}{2m\omega} \left(2n+1\right)}\\
%%   \Delta p &= \sqrt{<p^2>-<p>^2}\\
%%   &= \sqrt{\frac{\hbar m\omega}{2}\left(2n + 1\right)}\\
%%   \Delta x \Delta p &= \sqrt{\frac{\hbar}{2m\omega} \left(2n+1\right)}
%%   \sqrt{\frac{\hbar m\omega}{2}\left(2n + 1\right)}\\
%%   &= \frac{\hbar}{2}(2n+1)\\
%% \end{align*}
%% This is an interesting result which indicates that uncertainty increases as the pure wave function
%% gets more energetic. Ground state is the state with maximum certainty. The higher states get
%% progressively more uncertain, which makes sense since both the velocity and the well get larger
%% for higher states.\\

%% \section{9.14: Unknown QHO wavefunction}
%% We are given the wave function $\psi(x,0) = A\left[1 - 3\sqrt{\frac{m\omega}{\hbar}}x +
%%   2\frac{m\omega}{\hbar}x^2\right]$. Because its highest-order polynomial term is $x^2$, we know
%% that it must be made up of the $n = 0, 1, 2$ energy states. We now figure out the coefficients for
%% each:

%% \begin{align*}
%%   c_0 &= \int_{-\infty}^{\infty} \phi_0^*\psi dx\\
%%   &= A\int_{-\infty}^{\infty} \left(\frac{m\omega}{\pi\hbar}\right)^{1/2}e^{-m\omega x^2/\hbar}
%%   \left(1 - 3\sqrt{\frac{m\omega}{\hbar}}x +
%%   2\frac{m\omega}{\hbar}x^2\right)e^{-m\omega x^2/(2\hbar)}dx\\
%%   &= 2\\
%%   c_1 &= A\int_{-\infty}^{\infty} \frac{2\sqrt{\frac{m\omega}{\hbar}x}}{\sqrt{2}}
%%     \left(\frac{m\omega}{\pi\hbar}\right)^{1/2} \left(1 - 3\sqrt{\frac{m\omega}{\hbar}}x +
%%     2\frac{m\omega}{\hbar}x^2\right)e^{-m\omega x^2/\hbar}dx\\
%%  &= -\frac{3}{\sqrt{2}}\\
%%  c_2 &= A\int_{-\infty}^{\infty} \frac{\frac{4m\omega}{\hbar}x^2 - 2}{\sqrt{8}}
%%     \left(\frac{m\omega}{\pi\hbar}\right)^{1/2} \left(1 - 3\sqrt{\frac{m\omega}{\hbar}}x +
%%     2\frac{m\omega}{\hbar}x^2\right)e^{-m\omega x^2/\hbar}dx\\
%%  &= \sqrt{2}\\
%% \end{align*}

%% From this we can see that $\psi(x,0) = \sqrt{\frac{21}{2}}
%% \left(2\ket{0} + -\frac{3}{\sqrt{2}}\ket{1} + \sqrt{2}\ket{2}\right)$. It is then easy to compute
%% the expectation value: \\

%% \begin{align*}
%%   <E> &= \sigma \hbar\omega\left(n+\frac{1}{2}\right)|c_n|^2\\
%%   &= \frac{2 + \frac{27}{4} + 5}{\frac{21}{2}} = 1.3 eV\\
%% \end{align*}

\section{6: Position, momentum commutator}
Here we find the value of the position, momentum commutator:
\begin{align*}
  [\hat{x}, \hat{p}] &= \hat{x}\hat{p} - \hat{p}\hat{x}\\
  &= \left(\hat{x}\hat{p}\psi(x) - \hat{p}\hat{x}\psi(x)\right)\frac{1}{\psi(x)}\\
  &= \left(\hat{x}\left(-i\hbar\frac{d}{dx}\right)\psi(x) - \hat{p}x\psi(x)\right)\frac{1}{\psi(x)}\\
  &= \left(-i\hbar x \frac{d\psi}{dx} + i\hbar\left(\psi(x)
  + x\frac{d\psi}{dx}\right)\right)\frac{1}{\psi(x)}\\
  &= \left(i\hbar\psi(x)\right)\frac{1}{\psi(x)}\\
  &= i\hbar\\
\end{align*}
Here we use the definitions of the position and momentum operators to find the value of their
commutator. We need to operate on a generic wave function in order to do algebra on the
operators, since the differential in the momentum operator is meaningless without a function
to act on. This result is critical, since it is at the heart of the Heisenberg Uncertainty
Principle: position and momentum don't commute, so they do not share a basis and they are not
co-measurable. This is the mathematical representation of the popular saying ``you cannot know
the position and momentum of a particle simultaneously.''

\section{7: Commutator algebra}

\begin{align*}
  \left[\hat{A},\hat{A}^n\right] &= AA^n - A^nA = A^{n+1} - A^{n+1} = 0\\
  \left[\hat{A},\hat{B}\right] &= \hat{A}\hat{B} - \hat{B}\hat{A}
  = -\left(\hat{B}\hat{A} - \hat{A}\hat{B}\right) = -\left[\hat{B},\hat{A}\right]\\
  \left[\hat{A},c\hat{B}\right] &= \hat{A}c\hat{B} - c\hat{B}\hat{A}
  = c\left(\hat{A}\hat{B} - \hat{B}\hat{A}\right) = c\left[\hat{A},\hat{B}\right]\\
  \left[\hat{A},\left(\hat{B}+\hat{C}\right)\right] &= \hat{A}\left(\hat{B} + \hat{C}\right)
  - \left(\hat{B} + \hat{C}\right)\hat{A} = \hat{A}\hat{B} - \hat{B}\hat{A}
  + \hat{A}\hat{C} + \hat{C}\hat{A} = \left[\hat{A},\hat{B}\right] + \left[\hat{A},\hat{C}\right]\\
  \left[\hat{A},\hat{B}\hat{C}\right] &= \hat{A}\hat{B}\hat{C} - \hat{B}\hat{C}\hat{A}
  = \hat{A}\hat{B}\hat{C} - \hat{B}\hat{A}\hat{C} + \hat{B}\hat{A}\hat{C} - \hat{B}\hat{C}\hat{A}
  = \left[\hat{A},\hat{B}\right]\hat{C} + \hat{B}\left[\hat{A},\hat{C}\right]\\
  \left[\hat{A},\left[\hat{B},\hat{C}\right]\right]
  + \left[\hat{B},\left[\hat{C},\hat{A}\right]\right]
  &+ \left[\hat{C},\left[\hat{A},\hat{B}\right]\right] =
  \left[\hat{A},\hat{B}\hat{C} - \hat{C}\hat{B}\right]
  + \left[\hat{B},\hat{C}\hat{A} - \hat{A}\hat{C}\right]
  + \left[\hat{C},\hat{A}\hat{B} - \hat{B}\hat{A}\right] = \\
  &= \hat{A}\hat{B}\hat{C} - \hat{A}\hat{C}\hat{B}
  + \hat{B}\hat{C}\hat{A} - \hat{B}\hat{A}\hat{C}
  + \hat{C}\hat{A}\hat{B} - \hat{C}\hat{B}\hat{A}\\
\end{align*}
